---
title: "Boxplot for Gene"
subtitle: ""
author: "Ximing Ran"
date: "2025-03-31"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: false
  html_document:
    # code_folding: hide
    toc: true
    toc_float: true
hitheme: tomorrow
highlighter: highlight.js

---

```{r setup, include=FALSE}
# load libraries
library(tidyverse)
library(knitr)
set.seed(2024)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "./results/04-gene-boxplot/Analysis_figure/"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
knitr::kable(head(mtcars[, 1:4]), "simple")
```

```{r}
library(tibble)
library(tidyr)
library(dplyr)
library(rtracklayer)
```



```{r local_function_load}
# load function from local files
source(here::here("source", "DEG_functions.R"))
```

\newpage

# 1. Read the count data
In this section, we will read the clean count data from the synaptosomes_miRNARNA folder. The data is stored in the format of `.csv`. We will read the data and filter the columns based on the condition list. The final table will be stored in `results/02-DEG-Vehicle/DESeq2_results.csv`.

```{r load_bulkRNA_data}
input_count <- read.csv(here::here("data", "synaptosomes_bulkRNA",
                                      "synaptosomes_bulkRNA_counts_cleaned.csv"))
counts <-  as.data.frame(input_count) %>% 
  column_to_rownames(var = "gene")
colnames(counts) <- gsub("_", "-", colnames(counts))


# raw sample list
sample_list_raw <- read.csv(here::here("data", "synaptosomes_bulkRNA",
                                      "sample_info_22q.csv")) %>%
                    mutate(condition = paste0(Diagosis, "_", Treatment))


# Ensure the column names of counts exist in Sample.name
new_colnames <- sample_list_raw$Label[match(colnames(counts), sample_list_raw$Sample.name)]

# Assign new column names
colnames(counts) <- new_colnames


# sort the columns by the colname
condition_list <- data.frame(
  group =sample_list_raw$condition
)

row.names(condition_list) <- sample_list_raw$Label

counts<- counts[, rownames(condition_list)]

# load the 22q gene_list
target_22q_gene <- read.csv(here::here("data","ref" ,"22q_gene_2024_10_17.csv"))
target_gene <- target_22q_gene$gene
target_gene <- target_gene[1:(length(target_gene) - 4)]

gene_name_mapping<- readRDS(here::here("data","ref" ,"gene_name_mapping.rds"))




```


# 2. Normalization and DESeq2 analysis

```{r DESeq2_analysis}
# filter gene withall 0
count_input <-  counts
count_input <- count_input[rowSums(count_input) > 0, ]

# Create a DESeq2 object
dds_obj <- DESeqDataSetFromMatrix(countData = count_input,
                              colData = condition_list,
                              design = ~ group)

# get the normalized count
dds_obj <- DESeq(dds_obj)

# Get the normalized count
vsd.obj <- varianceStabilizingTransformation(dds_obj, blind = TRUE)
normalized_count  <- assay(vsd.obj)


# save the normalized count to the file
write.csv(normalized_count , file = file.path("results","04-gene-boxplot", "DESeq2_normalized_count.csv"))


```


\newpage

# 3. Visualization for reuslt

## (1) Sample information Heatmap 

```{r Sample_dis_Heatmap, fig.width=8, fig.height=6}
result_folder = file.path("results", "04-gene-boxplot")
# Plot the sample distance 
plot_sample_heatmap(dds_obj, figure_folder =  file.path(result_folder,"01-Sample_info"),
                    save = TRUE,
                    file_name = "01_sample_distance_heatmap")
```

\newpage

## (2) Sample information PCA plot

```{r  Sample_PCA, fig.width=8, fig.height=6}
# Plot the PCA plot for the sample
plot_sample_PCA_plot_mutil(dds_obj,figure_folder = file.path(result_folder,"01-Sample_info"),
                   file_name = "02_sample_PCA_plot")
```

\newpage

```{r boxplot_gene, fig.width=4, fig.height=6}
counts_matrix <- counts(dds_obj, normalized = TRUE)


condition_list_label <- condition_list 
condition_list_label$group <- gsub("_", "+", condition_list_label$group)
condition_list_label$group <- gsub("Vehicle", "Veh", condition_list_label$group)

all_gene <- target_gene 


for (index in 1:length(all_gene )){
  plot_gene<- all_gene [index]
  
  p<-plot_expression_boxplot_4_group(counts_matrix = normalized_count, condition_list_label, feature =  plot_gene ,
                                            figure_folder = file.path(result_folder,"Boxplot"),
                                             file_name = paste0("gene_", plot_gene),
                                              fig.height = 6, fig.width = 4,
                                              save=TRUE,
                                              group_1="CTRL+Veh", group_2="CTRL+TTX",
                                              group_3="22q+Veh", group_4="22q+TTX")
  print(p)
  print(paste0("gene_", plot_gene))
}


```


\newpage
# Session information
```{r}
sessionInfo()
```

